Unified visualization of seasonal influenza evolution in the laboratory and in nature.

### 2019-06-11 paper/paper.tex
Add title declaration back to head of paper
This fixes the two column layout.
Adhere to one sentence per line standard
Makes tex files easier to read on GitHub and easier to manage with git (changing
one word in a sentence doesn't appear to change an entire paragraph). Also
replaced unicode apostrophes with ASCII versions and escaped percentage symbols.

@@

Seasonal influenza is an infectious disease that infects between 10 and 20% of the world’s population every year \cite{neher2015nextflu}. For some diseases, infection confers immunity that protects that person from subsequent infections for many years. In other cases, such as with influenza virus, the surface proteins on the pathogen change quickly, thereby escaping recognition by antibodies elicited by the previous infection. It is for this reason that the influenza vaccine must be updated almost annually, since the immune response elicited by  the vaccine needs to match whichever viruses we expect to circulate during the next influenza season. Vaccination remains the primary public health response effort to control the spread of influenza.

@@

\section{Related Work}
The Bloom lab at Fred Hutch performs are large amount of DMS work to investigate functional stability and immune escape and selection for important human pathogens such as HIV, influenza, and Zika virus. Originally developed by Thyagarajan and Bloom \cite{thyagarajan2014inherent}, initial DMS experiments on influenza virus hemagglutinin focused on understanding which sites of the protein could tolerate multiple amino acids and cause a functional infection. These studies have been extended to include not only purifying selective forces, but also immune selection as well, by innate immunity \cite{ashenberg2017deep} and adaptive immunity involving antibody response \cite{Lee2018}. In total, these experiments have helped us to define broad atlases of paths for escaping immunity \cite{dingens2019antigenic}. Importantly, the development of such atlases, and the visualization of these data, requires an intensive computational workflow that involves iteratively generating static graphs for the relevant streams of data: immune selection as measured by the DMS experiment and the 3D structure of the protein. By default these images also do not link a third relevant data stream, and that is the frequencies of amino acids observed at different sites of the protein in nature. Thus the task of our visualization platform is to provide an easy way to view and interact with these three data streams in a linked fashion that is also easy to use in the absence of coding expertise.

\section{Methods}
This visualization platform links three data views together by their common shared element, site in the protein. The three data streams are: 1) DMS differential selection value, which is a quantitative measure of the preference for a certain amino acid at a specific site in the protein, 2) the natural amino acid frequencies of a site, a quantitative variable the describes what proportion of all sampled strains of influenza at time T have a specific amino acid at a site, and 3) the 3-D structure of protein itself, which takes linear sequence space and folds that up to reflect where certain sites of the sequence fall on the protein structure. We describe design decisions about how to visualize these data streams below. Our visualization was implemented using D3 \cite{bostock2011d3}

@@

\subsection{DMS differential selection preferences}
Navigation of our visualization begins with the user selecting a site in the gene that they would like to explore. Because the gene segments can be long, and many measurements are highly similar, it would be challenging to observe all measurements at once and interact with them. To address this challenge, we have two plots within this data stream’s panel, one that gives context about the overall positioning in the gene segment, and one that is the zoomed view that can be interacted with.  To zoom in, the context panel can be brushed, and this will select the portion of the gene segment that the user would like to zoom in on. Within the zoom panel, the DMS preferences by site, encoded using position, can be interacted with by mousing over the data points. Mousing over brings up a tooltip providing additional details about the DMS data at that site. In particular, the tooltip gives the exact value of the differential selection measurement, since this might be hard to read from looking at the y-axis. The points can also be selected by clicking on them, which allows the user to enter the linked visualization, whereby the natural amino acid frequencies for that site, and that site in the 3D protein structure, will be selected.

\begin{figure}[H]
	\includegraphics[width=1.0\textwidth]{dms-brushed-tooltip.png}
	\caption{\textbf{Visualization of differential selection data generated from DMS experiments.}
  The laboratory data can be explored and zoomed in on by brushing the context panel below the primary panel, giving both detail and overview to the user. Sites that are colored black in this panel represent sites for which there is natural frequency data. For all grey sites, only one variant of influenza at that site is found to circulate naturally. Mousing over a tip brings up details on demand, and clicking on a data point selects that site for visualization in the other panels, and colors the point according to it's maximal differential selection value.
	}
	\label{dms-data}
\end{figure}

@@

\begin{figure}[H]
	\includegraphics[width=1.0\textwidth]{3d-structure-panel.png}
	\caption{\textbf{3D protein structure with selected site highlighted.}
  3D protein structure panel, with the selected site colored according to it's differential selection value. The tooltip also provides information about which amino acid is at highest frequency at the site. In this image, site 193 has a PHE, or phenylalanine.
  }
	\label{3d-structure}
\end{figure}

@@

DMS experiments are conducted for various different genes, different pathogens, and outside of the field of infectious disease research as well. As such, in the future we would like to expand this visualization platform to accept data sets other than from influenza. To facilitate wider use, we would like to build functionality into the visualization platform such that it can take in drag-and-drop files for other protein structures and preferences. Additionally, we would like to expand the number of panels that would allow joint visualization, for instance, by including phylogenetic trees that would allow querying of specific viruses with linking to relevant information about the virus and it’s proteins to relevant DMS data.

>>>>

\maketitle

@@

Seasonal influenza is an infectious disease that infects between 10 and 20\% of the world's population every year \cite{neher2015nextflu}.
For some diseases, infection confers immunity that protects that person from subsequent infections for many years.
In other cases, such as with influenza virus, the surface proteins on the pathogen change quickly, thereby escaping recognition by antibodies elicited by the previous infection.
It is for this reason that the influenza vaccine must be updated almost annually, since the immune response elicited by  the vaccine needs to match whichever viruses we expect to circulate during the next influenza season.
Vaccination remains the primary public health response effort to control the spread of influenza.

@@

\section{Related Work}

The Bloom lab at Fred Hutch performs are large amount of DMS work to investigate functional stability and immune escape and selection for important human pathogens such as HIV, influenza, and Zika virus.
Originally developed by Thyagarajan and Bloom \cite{thyagarajan2014inherent}, initial DMS experiments on influenza virus hemagglutinin focused on understanding which sites of the protein could tolerate multiple amino acids and cause a functional infection.
These studies have been extended to include not only purifying selective forces, but also immune selection as well, by innate immunity \cite{ashenberg2017deep} and adaptive immunity involving antibody response \cite{Lee2018}.
In total, these experiments have helped us to define broad atlases of paths for escaping immunity \cite{dingens2019antigenic}.
Importantly, the development of such atlases, and the visualization of these data, requires an intensive computational workflow that involves iteratively generating static graphs for the relevant streams of data: immune selection as measured by the DMS experiment and the 3D structure of the protein.
By default these images also do not link a third relevant data stream, and that is the frequencies of amino acids observed at different sites of the protein in nature.
Thus the task of our visualization platform is to provide an easy way to view and interact with these three data streams in a linked fashion that is also easy to use in the absence of coding expertise.

\section{Methods}

This visualization platform links three data views together by their common shared element, site in the protein.
The three data streams are: 1) DMS differential selection value, which is a quantitative measure of the preference for a certain amino acid at a specific site in the protein, 2) the natural amino acid frequencies of a site, a quantitative variable the describes what proportion of all sampled strains of influenza at time T have a specific amino acid at a site, and 3) the 3-D structure of protein itself, which takes linear sequence space and folds that up to reflect where certain sites of the sequence fall on the protein structure.
We describe design decisions about how to visualize these data streams below.
Our visualization was implemented using D3 \cite{bostock2011d3}

@@

\subsection{DMS differential selection preferences}
Navigation of our visualization begins with the user selecting a site in the gene that they would like to explore.
Because the gene segments can be long, and many measurements are highly similar, it would be challenging to observe all measurements at once and interact with them.
To address this challenge, we have two plots within this data stream's panel, one that gives context about the overall positioning in the gene segment, and one that is the zoomed view that can be interacted with.
 To zoom in, the context panel can be brushed, and this will select the portion of the gene segment that the user would like to zoom in on.
Within the zoom panel, the DMS preferences by site, encoded using position, can be interacted with by mousing over the data points.
Mousing over brings up a tooltip providing additional details about the DMS data at that site.
In particular, the tooltip gives the exact value of the differential selection measurement, since this might be hard to read from looking at the y-axis.
The points can also be selected by clicking on them, which allows the user to enter the linked visualization, whereby the natural amino acid frequencies for that site, and that site in the 3D protein structure, will be selected.

\begin{figure}[H]
	\includegraphics[width=1.0\textwidth]{dms-brushed-tooltip.png}
	\caption{\textbf{Visualization of differential selection data generated from DMS experiments.}
  The laboratory data can be explored and zoomed in on by brushing the context panel below the primary panel, giving both detail and overview to the user.
Sites that are colored black in this panel represent sites for which there is natural frequency data.
For all grey sites, only one variant of influenza at that site is found to circulate naturally.
Mousing over a tip brings up details on demand, and clicking on a data point selects that site for visualization in the other panels, and colors the point according to its maximal differential selection value.
	}
	\label{dms-data}
\end{figure}

@@

\begin{figure}[H]
	\includegraphics[width=1.0\textwidth]{3d-structure-panel.png}
	\caption{\textbf{3D protein structure with selected site highlighted.}
  3D protein structure panel, with the selected site colored according to its differential selection value.
The tooltip also provides information about which amino acid is at highest frequency at the site.
In this image, site 193 has a PHE, or phenylalanine.
  }
	\label{3d-structure}
\end{figure}

@@

DMS experiments are conducted for various different genes, different pathogens, and outside of the field of infectious disease research as well.
As such, in the future we would like to expand this visualization platform to accept data sets other than from influenza.
To facilitate wider use, we would like to build functionality into the visualization platform such that it can take in drag-and-drop files for other protein structures and preferences.
Additionally, we would like to expand the number of panels that would allow joint visualization, for instance, by including phylogenetic trees that would allow querying of specific viruses with linking to relevant information about the virus and its proteins to relevant DMS data.
